Grove: A Bidirectionally Typed Collaborative Structure Editor Calculus

### 2024-10-25 discussion-and-conclusion.tex
updates for final submission

\section{Discussion and Conclusion}%
\label{sec:Discussion and Conclusion}

TODO: leaves

TODO: our model supports treating these as a cons-list of characters

TODO: (Place somewhere) This move semantics gives us a richer structure than
when treating code as a list of lines.  We exploit
this in Section~REF:TODO in order merge edits that involve moving code.

TODO: Traditional diff: no relation (and indent might change)

TODO: memory usage

TODO: cache eviction algorithm

TODO: Finally, note that through we present several complex scenarios, this is merely for presentation.
In practice, these complex scenarios occur less frequently than portrayed here.

\subsection{Variable names, strings, and numbers}%
\label{sub:Variable names, strings, and numbers}

NOTE: we could implement each digit as a separate characters

GUI for string conflicts: use popups

>>>>

\section{Discussion and Conclusion}%
\label{sec:Discussion and Conclusion}
\begin{quote}
    \textit{``The fact that commutation can fail [in Darcs] makes a huge difference in the whole patch formalism. It may be possible to create a formalism in which commutation always succeeds, with the result of what would otherwise be a commutation that fails being something like a virtual particle ... and it may be that such a formalism would allow strict mathematical proofs ... However, I’m not sure how you’d deal with a request to delete a file that has not yet been created, for example. Obviously you’d need to create some kind of antifile, which would annihilate with the file when that file finally got created ...''}

    -- David Roundy, Theory of patches~\cite{old-darcs-manual}
\end{quote}

This paper proposes a radically simpler, albeit practically ambitious, rearchitecture of collaborative editing. Our contributions together result in a typed collaborative structure calculus called Grove 
where, uniquely, all edits, including code relocations that stymie existing approaches, commute and where there are no semantic gaps: all possible editor states, including editor states with various kinds of unresolved conflicts, are semantically meaningful.

This paper focuses on the core theoretical underpinnings of this approach,
developing mechanized metatheory for both the patch language and the type system. 
A number of research problems on algorithmic, networking, and user interface aspects of the problem open up given these foundations. For example, we point out several situations where presenting non-conflicted but heuristically attention-worthy merges may be worthwhile, and we leave to future work the user experience design of this process. We intend to use the Grove workbench to integrate these efforts into the Hazel programming environment, though its editor component has been evolving so rapidly as to prevent experimentation in this direction so far.

Although our focus was on tree editing, some aspects of a program are more naturally linearly structured, e.g. string literals. We also leave to future work the problem of combining existing work on sequence CRDTs~\cite{ahmed2011evaluating,kleppmann2020moving} with our work on tree/graph CRDTs.

% In practice, it may be helpful to garbage collect orphaned vertices once there is consensus across collaborators that the deletion is permanent, but we do not consider this consensus protocol formally in this paper. In an open-ended collaboration scenario (where the set of collaborators is not known, e.g. on GitHub), we simply retain all vertices.

% TODO: talk about linear sequences in code

% TODO: leaves

% TODO: our model supports treating these as a cons-list of characters

% TODO: (Place somewhere) This move semantics gives us a richer structure than
% when treating code as a list of lines.  We exploit
% this in Section~REF:TODO in order merge edits that involve moving code.

% TODO: Traditional diff: no relation (and indent might change)

% TODO: memory usage

% TODO: cache eviction algorithm

% TODO: Finally, note that through we present several complex scenarios, this is merely for presentation.
% In practice, these complex scenarios occur less frequently than portrayed here.

% \subsection{Variable names, strings, and numbers}%
% \label{sub:Variable names, strings, and numbers}

% NOTE: we could implement each digit as a separate characters

% GUI for string conflicts: use popups

### 2024-10-25 grove-paper.tex
updates for final submission

\documentclass[acmsmall,dvipsnames,10pt,review,anonymous,nonacm]{acmart}\settopmatter{printfolios=true}

@@

% figures
\input{figures/figure-simple}
\input{figures/figure-wrap-move}
\input{figures/figure-different-parts-nested-parts}
\input{figures/figure-multi-child}
\input{figures/figure-multi-parent}
\input{figures/figure-cycle}
\input{figures/figure-disconnect}
\input{figures/figure-decomposition}
\input{figures/figure-term-syntax}
\input{figures/figure-arity}
\input{figures/figure-decomposition-def}
\input{figures/figure-user-actions}

@@

%% Author information
\author{Michael D. Adams}
\orcid{0000-0003-3160-6972}

\author{Eric Griffis}
\orcid{0000-0003-1693-6172}

\author{Sundara Vishnu Satish}
\orcid{TODO}

\author{Thomas J. Porter}
\orcid{TODO}

\author{Cyrus Omar}
\orcid{0000-0003-4502-7971}
\affiliation{
  %\position{Assistant Research Scientist}
  \department[0]{Computer Science and Engineering}
  \department[1]{Electrical Engineering and Computer Science}
  \department[2]{College of Engineering}
  \institution{University of Michigan}
  \streetaddress{Bob and Betty Beyster Building, 2260 Hayward Street}
  \city{Ann Arbor}
  \state{MI}
  \postcode{48109-2121}
  \country{USA}
}

@@

\input{introduction}
\input{grove-by-example}
\input{formalism}
\input{marking-system}
\input{implementation}
\input{related-work}
\input{discussion-and-conclusion}

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
  %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Appendix
\appendix

@@

%% Temporary appendixes that will not be published go **AFTER** the bibliography and do not count towards the page count

\clearpage % Put these appendixes on separate pages so we can easily remove them from the document.  Also ensure all figures have been placed.

% \input{appendix-A-graph-sequences}
\input{appendix/appendix-A-formalism}  
% \input{appendix/appendix-B-marking-system}
\input{appendix/appendix-B-typechecking}
% \input{appendix/appendix-C-type-inference}

>>>>

\documentclass[acmsmall,dvipsnames,10pt,review]{acmart}\settopmatter{printfolios=true}
% Need to tweak for camera-ready submission
% \documentclass[acmsmall,dvipsnames,10pt]{acmart}\settopmatter{printfolios=true}

@@

% figures
\input{figures/figure-simple}
\input{figures/figure-wrap-move}
\input{figures/figure-different-parts-nested-parts}
\input{figures/figure-multi-child}
\input{figures/figure-multi-parent}
\input{figures/figure-cycle}
\input{figures/figure-disconnect}
\input{figures/figure-decomposition}
\input{figures/figure-term-syntax}
\input{figures/figure-arity}
\input{figures/figure-decomposition-def}
\input{figures/figure-user-actions}
\input{figures/figure-user-action-transitions}

@@

%% Author information
\author{Michael D. Adams}
\orcid{0000-0003-3160-6972}
\affiliation{
\institution{National University of Singapore}
\country{Singapore}
}

\author{Eric Griffis}
\orcid{0000-0003-1693-6172}
\affiliation{
\institution{University of Michigan}
\city{Ann Arbor}
\state{MI}
\country{USA}
}

\author{Thomas J. Porter}
\orcid{0009-0000-1809-8382}
\affiliation{
\institution{University of Michigan}
\city{Ann Arbor}
\state{MI}
\country{USA}
}

\author{Sundara Vishnu Satish}
\orcid{0009-0002-8685-1342}
\affiliation{
\institution{University of Michigan}
\city{Ann Arbor}
\state{MI}
\country{USA}
}

\author{Eric Zhao}
\orcid{0009-0000-4969-2376}
\affiliation{
\institution{University of Michigan}
\city{Ann Arbor}
\state{MI}
\country{USA}
}

\author{Cyrus Omar}
\orcid{0000-0003-4502-7971}
\affiliation{
  %\position{Assistant Research Scientist}
  % \department[0]{Computer Science and Engineering}
  % \department[1]{Electrical Engineering and Computer Science}
  % \department[2]{College of Engineering}
  \institution{University of Michigan}
  % \streetaddress{Bob and Betty Beyster Building, 2260 Hayward Street}
  \city{Ann Arbor}
  \state{MI}
  % \postcode{48109-2121}
  \country{USA}
}

@@

\input{introduction}
\input{grove-by-example}
\input{formalism}
\input{implementation}
\input{marking-system}
\input{related-work}
\input{discussion-and-conclusion}

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
  %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.

  This work was partially funded by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{2238744}. Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Appendix
% \appendix

@@

### 2024-10-25 related-work.tex
updates for final submission

Contrast with a traditional diff

Similar to how git merged across moves.
A notable difference though, is that \texttt{git} detects
file moves by structural similarity rather than
our nominal equality.
(We exploit this in Subsection~REF:TODO).

Compare this to what happens when there are conflicts in a traditional,
line/diff-based version control system (e.g., \texttt{git}, \texttt{mercurial}, \texttt{svn}, etc.).
Those systems annotate files with difference markers (e.g., sequences of~\texttt{<},~\texttt{>}, or~\texttt{=})
when a merge conflict happens.
These show alternate versions of the code, and it is up to the user
to replace those alternates with the desired merged result.

Another place our model is different
is that changes are explicit rather than being inferred by a \texttt{diff} algorithm.
Finally, in our model, since all changes are in terms of
the structure of the code rather than lines,
all changes respect that structure.
This is unlike the line-based model where there is no guarantee
that a merge conflict follows the grammatical structure of the code.

\subsection{Trees}

\subsubsection{Structure editors}

Hazel~\citep{Omar:2019:10.1145/3290327}

\subsubsection{Tree diff}

>>>>

The core components of a version control system are a patch language, a method for synthesizing patches from user actions, and an approach for merging patches. 
% Existing patch based systems such as Darcs~\cite{DBLP:conf/haskell/Roundy05} and Pijul achieve commutativity between patches in many cases, but Grove achieves total commutativity.

% Grove stands out from existing systems in the precision of its patch language and the simplicity of its patch synthesis and merge processes, as well as its intrinsic guarantee of commutativity by employing a graph CmRDT as its repository data structure.

% Patch languages vary in their native data structure and level of precision, with Grove's patch language being exceptionally fine-grained and molded to the language. Traditional patch synthesis algorithms reconstruct patches from version differences, while Grove uniquely synthesizes patches directly from edit actions. Traditional merge algorithms, such as those used by Git, employ operational transforms to

% Traditional patch synthesis and merge algorithms, such as those used by Git, are complex and indirect, while Grove joins a family of new systems employing

\subsection{Patch Languages}
There have been many different designs for patch languages and indeed many imperative data structures and their associated operations can be construed as patch languages in the most general sense. In the context of collaborative coding, patch languages can differ in the data structures they operate on (e.g. line-based text~\cite{DBLP:conf/haskell/Roundy05}, character-based text~\cite{DBLP:conf/sigmod/EllisG89,DBLP:journals/pacmhci/LittLKH22}, tree-structured data). 
They can also differ in how they identify locations within the data (e.g. by using numeric offsets~\cite{DBLP:conf/sigmod/EllisG89}, unique identifiers, or paths through a tree).
Finally, patch languages differ in which specific actions are supported explicitly. Insertion and deletion are common, while code relocation, copying, undo, and other operations are variously also included.

In this paper, our focus was on syntax trees with holes (i.e. program sketches~\cite{DBLP:conf/aplas/Solar-Lezama09, DBLP:conf/popl/OmarVHAH17}) and explicit conflicts, which we represented as directed graphs. We identify locations using unique IDs. We have a two-level patch language, with a low-level graph patch language supporting only edge insertion and deletion and a higher-level user edit action language focused on insertion, deletion, and relocation, with some additional narrative consideration of copying and undo (a fuller account of which we leave to future work).  Our user edit language therefore forms a structure editor calculus, inspired closely by recent work on the Hazelnut structure editor calculus (which did not support relocation)~\cite{DBLP:conf/popl/OmarVHAH17} and patch languages for other tree-based data structures~\cite{DBLP:conf/sigmod/ChawatheG97, DBLP:journals/tse/FluriWPG07,DBLP:conf/kbse/FalleriMBMM14,DBLP:conf/doceng/Lindholm04,DBLP:conf/fase/NguyenNPN10,DBLP:journals/scp/SchwagerlUW15}.

\subsection{Patch Synthesis}
The most common approach to patch synthesis is to use a differencing algorithm to compare two states, e.g. from the file system, to generate a patch. 
The classic \li{diff} algorithm~\cite{DiffAlgorithm}, for example, minimizes edit distance for a patch language involving line insertions and deletions.

In the setting of tree-based editing, there have been a number of tree differencing algorithms described in the literature \cite{DBLP:conf/esa/Klein98,DBLP:journals/tcs/Bille05,DBLP:journals/talg/DemaineMRW09,DBLP:journals/fuin/AratsuHK10,DBLP:conf/sigmod/ChawatheG97, DBLP:journals/tse/FluriWPG07,DBLP:conf/kbse/FalleriMBMM14,DBLP:conf/doceng/Lindholm04,DBLP:conf/fase/NguyenNPN10,DBLP:journals/scp/SchwagerlUW15}. 
As described in \autoref{sec:Introduction}, synthesizing insertions and deletions is well-understood, but synthesizing relocations is more complex and requires heuristics.

The approach we explore in this paper is far simpler: we directly translate from the log of user edit actions to graph patches, without needing a differencing algorithm at all. This is only possible with a structure editor integrated with the version control system, but the benefit of direct visibility into the edit action log is that we do not need heuristics to synthesize relocations.

\subsection{Merging}
\subsubsection{Operational Transforms}
The most common approach to merging concurrently developed patches is to deploy an operational transform~\cite{DBLP:conf/sigmod/EllisG89} whereby locations in a remote patch are modified based on the action of a local patch. Standard three-way merge algorithms in version control systems deploy this approach, as do real-time collaborative editors. There have been a number of papers studying the algebraic properties of merging patches in this style.

For example, the Darcs~\cite{DBLP:conf/haskell/Roundy05} version control system, like Grove, represents repositories using sets of patches. Using operational transforms, Darcs achieves commutativity in many cases, but not between conflicting patches. The theory of Darcs defines and algebraically characterizes when operational transforms do satisfy the properties of associativity and commutativity. Recent work on homotopical patch theory \cite{DBLP:journals/jfp/AngiuliMLH16} has similarly developed an abstract algebraic framework for distinguishing sensible merges.

\subsubsection{CRDTs}

The observation that commutativity is a particularly helpful property when dealing with concurrent systems, including version control systems, has led to the development of a number of data structures centered on commutativity. These are known as CRDTs, which stands variously for \emph{conflict-free}, \emph{convergent} (CvRDT), or \emph{commutative replicated datatypes} (CmRDT) depending on particular details about the operations and the state representation~\cite{shapiro2011conflict}.
Our approach draws directly from this line of work: the Grove patch language forms a CmRDT on directed graphs, from which we observe that we can derive a CmRDT for trees with explicit conflicts.

Other examples of CRDT-based editors include and Peritext~\cite{DBLP:journals/pacmhci/LittLKH22} and Zed~\cite{zed-blog}. These systems differ from Grove by targeting synchronous editing rather than version control and by operating on sequential text rather than tree structures.

There have been other recent efforts to develop CmRDTs for tree data structures~\cite{DBLP:conf/icdcs/PreguicaMSL09, Highly-Available}. However, the focus in this work has been on avoiding conflicts and cycles entirely by applying \emph{ad hoc} heuristics for conflict resolution at merge-time, e.g. using reported timestamps or favoring particular directions in the tree. Our approach instead embraces manual conflict resolution, as is common practice in software projects where arbitrarily losing code is not acceptable.

% Zed is a Integrated Development Environment (IDE) that reasons text edits using CRDTs and differs in approach by using Lamport timestamps to order events across replicas. Peritext, on the other hand, is a character based CRDT editor but is limited to the realm of rich-text editing.

Pijul iterates on the patch based system of Darcs, obtaining commutativity between conflicting patches using a CRDT graph data structure~\cite{pijul-talk}. Pijul's graph data structure is very similar to that of Grove in its treatment of edges, but, unlike Grove, requires vertices to be created before they can be referred to. This imposes a dependency relation between patches, causing Pijul to fall short of the full commutativity enjoyed by Grove.
Pijul is also language agnostic and models only the linear structure of text. However, Pijul goes beyond this version of Grove by extending the data structure to support history and branches, and Grove may be similarly extended in future work.

% \url{https://www.waitingforcode.com/big-data-algorithms/conflict-free-replicated-data-types-flags-graphs-maps/read}
